Neural Co-Processors for Restoring Brain Function:\\Results from a Cortical Model of Grasping

### 2023-03-18 paper/v1.tex
v2 paper, after response to reviewers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\usepackage{graphicx,epstopdf}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{lscape}
\usepackage{hyperref}
\usepackage{microtype}
\graphicspath{ {./figs/} }
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\begin{document}

@@

% Authorship temporarily removed for double-anonymous review
\author{Matthew J Bryan$^{1}$, Linxing Preston Jiang$^{1,2,3}$, Rajesh P N Rao$^{1,2,3}$}

\address{$^{1}$ Neural Systems Laboratory, Paul G. Allen School of Computer
Science \& Engineering, University of Washington, Seattle, WA, USA}
\address{$^{2}$ Center for Neurotechnology, University of Washington, Seattle, WA, USA}
\address{$^{3}$ Computational Neuroscience Center, University of Washington, Seattle, WA, USA}

\ead{\{mmattb,prestonj,rao\}@cs.washington.edu}
\vspace{10pt}
\begin{indented}
\item[]March 2023
\end{indented}

\begin{abstract}
\textit{Objective} A major challenge in designing closed-loop brain-computer interfaces (BCIs)
is finding optimal stimulation patterns as a function of ongoing neural activity for
different subjects and different objectives. Traditional approaches, such as
those currently used for deep brain stimulation (DBS), have largely followed a manual
trial-and-error strategy to search for effective open-loop stimulation parameters,
a strategy that is inefficient and does not generalize to closed-loop
activity-dependent stimulation.
\textit{Approach} To achieve goal-directed closed-loop neurostimulation, we propose
the use of brain co-processors, devices which exploit artificial intelligence (AI)
to shape neural activity and bridge injured neural circuits for targeted repair and
restoration of function.  Here we investigate a specific type of co-processor called a ``neural
co-processor'' which uses artificial neural networks (ANNs) and deep learning to learn optimal
closed-loop stimulation policies. The co-processor adapts the stimulation policy as
the biological circuit itself adapts to the stimulation, achieving a form of brain-device
co-adaptation. Here we use simulations to lay the groundwork for future \textit{in vivo}
tests of neural co-processors. We leverage a previously published cortical model of grasping, to
which we applied various forms of simulated lesions. We used our simulations to develop
the critical learning algorithms and study adaptations to non-stationarity in preparation for
future \textit{in vivo} tests.
\textit{Main results} Our simulations show the ability of a neural co-processor to learn a
stimulation policy using a supervised learning approach, and to adapt that policy as
the underlying brain and sensors change. Our co-processor successfully co-adapted with
the simulated brain to accomplish the reach-and-grasp task after a variety of lesions
were applied, achieving recovery towards healthy function in the range 75-90\%.
\textit{Significance} Our results provide the first proof-of-concept demonstration,
using computer simulations, of a neural co-processor for adaptive activity-dependent closed-loop
neurostimulation for optimizing a rehabilitation goal after injury. While a significant
gap remains between simulations and \textit{in vivo} applications, our results provide
insights on how such co-processors may eventually be developed for learning complex
adaptive stimulation policies for a variety of neural rehabilitation and neuroprosthetic
applications. 
\end{abstract}

\vspace{2pc}
\noindent{\it Keywords}: brain-computer interface, brain-machine interface, neurostimulation,
neuromodulation, neural co-processor, AI, machine learning, deep learning, neural networks,
computational models
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

@@

The co-processor framework also allows co-adaptation with biological circuits in the brain by updating its 
stimulation policy, while the brain updates its own response to the stimulation via adaptation and neural
plasticity, or modifies its response due to other reasons. The co-processor could potentially optimize its outputs 
for a desired optimization function continually in the presence of significant non-stationarities in the brain.

Here, to lay the groundwork for future \textit{in vivo} tests of the co-processor framework, we 
use computer simulations to explore how co-processors can be trained to restore lost function and how they can adapt to 
 non-stationarities. We demonstrate a neural co-processor that restores movement in a
computational model of cortical networks involved in controlling a limb, after a simulated stroke affects the
ability to use that limb. Our demonstration combines both components of a neural co-processor \cite{rao.coproc}:
\begin{itemize}
	\item An emulation model based on ANNs, which learns a mapping from  
	      stimulation and current neural activity to output variables such as task performance (or future neural activity).
	\item An artificial intelligence (AI) ``agent'' based on ANNs which learns the best  closed-loop
          activity-dependent stimulation to apply in real time to optimize a given task.
\end{itemize}

\section{Background}
\label{sec:background}

Significant advances have been made in understanding and modeling the effects of electrical and other forms of
neural stimulation on the brain. Researchers have explored how information can be biomimetically or
artificially encoded and delivered via stimulation to neuronal networks in the brain and
other regions of the nervous system for auditory \cite{niparko.cochlear}, visual \cite{weiland.retinal},
proprioceptive \cite{tomlinson.propr}, and tactile
\cite{tabot.tact, tyler.tact, dadarlat.tact, sharlene.tact, cronin.tact} perception.
Advances have also been made in modeling the effects of stimulation over large scale, multi-region
networks, and across time \cite{shanechi.stimmodel}. Some models can additionally adapt to ongoing
changes in the brain, including changes due to the stimulation itself
\cite{tafazoli.acls}. For our simulations described below, we use a stimulation
model, not unlike those cited above, which seeks to account for both network dynamics
and non-stationarity. In addition to training the model to have a strong ability to predict
the effect of stimulation, we additionally adapt it to be useful for learning an
optimal stimulation policy, a property distinct from predictive
power alone.

Researchers have also explored both open- and closed-loop stimulation protocols for
treating a variety of disorders. Open loop stimulation has been effective in
treating Parkinson's Disease (PD) \cite{benabid.parkinsons}, as well as various
psychiatric disorders \cite{holtzheimer.psy, kisely.psy, fraint.psy}.
In research more directly related to our work, Khanna et al. \cite{khanna.openloop}
investigated the use of open loop stimulation in restoring dexterity after a lesion
in nonhuman primate's (NHP) motor cortex. The authors demonstrate that the use
of low-frequency alternating current, applied epidurally, can improve grasp performance.

While open loop stimulation techniques have yielded clinically useful results, results
in many domains have been mixed, such as in visual prostheses \cite{bosking.visual},
and in invoking somatosensory feedback \cite{cronin.tact}. We believe this is due to the
stimulation not being conditioned on the ongoing dynamics of the neural circuit being
stimulated. From moment to moment and throughout the day, a neuronal circuit in the brain
can be expected to respond differently even when the same stimulation parameters are used,
due to the multitude of different external and internal inputs influencing the circuit's
ongoing activity. Stimulation therefore needs to be closed-loop, i.e. proactively adapted
in response. This need is even greater over longer time scales as the effects of plasticity,
changes in clinical conditions, and ageing change the dynamics and connectivity of the brain.
Closed-loop stimulation may also provide means to better regulate the energy use of an
implanted stimulator, allowing it to intelligently regulate when to apply stimulation,
in order to preserve implant battery life \cite{castano.pd}. Another benefit is that closed-loop
stimulation offers an opportunity to minimize the side-effects of stimulation, through real
time regulation of the stimulation parameters, such as in the use of deep brain stimulation
(DBS) in PD patients \cite{little.park}. In recent years, closed-loop stimulation has been
used to aid in learning new memories after some impairment \cite{berger.closedloop,
kahana.biomarker}, to replay visually-invoked activations \cite{tafazoli.acls}, and for
optogenetic control of a thalamocortical circuit \cite{bolus.opto}, among others.

A major open question is: how does one leverage closed-loop stimulation for real-time
co-adaptation with the brain to accomplish an external task such as restoration of a lost function?
``Co-adaptation'' here refers to the ability of a BCI to adapt its stimulation regime to ongoing
changes in neural circuits in the brain, and to adapt with the brain to accomplish
the external task (e.g., grasping). The neural co-processor we present here provides one
potential approach to accomplishing this goal. Through the use of deep learning, a neural 
co-processor co-adapts its AI, which controls stimulation, in synch with the biological circuits in the brain.

For a neurologically complex task such as grasping, it is unlikely that there exists
a fixed real-time controller which can be identified \textit{a priori} for stimulating the (potentially
impaired) neural circuits involved in the task. This is due in large part to the variability
in the placement and performance of sensors and stimulators in different brains, as well as variability
in brain structure and function between subjects. The most plausible path to implementing a real-time
controller is therefore to allow the device to adapt to the subject, and to the long-term changes in
their brain activities, and variability in the sensors, stimulators and hardware. Our proposed neural
co-processors seek to accomplish such adaptation through ANNs and deep learning, and a particular
training paradigm described below.

\subsection{Simulation as a Way to Gain Insights Prior to {\em in vivo} Experiments}
To gain insights into neural co-processors before testing them \textit{in vivo}, we
investigated a number of crucial design elements through the use of a previously published model by
Michaels et al. \cite{michaels.mrnn} of the cortical areas involved in grasping; the model, based on
multiple recurrent neural networks, is inspired by cortical anatomy and was fit to data from nonhuman
primates performing grasping tasks. Using this cortical model as a ``simulated brain'' allowed us to
rapidly iterate through different design and training methods to demonstrate key properties of the
neural co-processor framework. These insights will help guide the co-processor training methodologies
and experimental design for future \textit{in vivo} experiments. Additionally, a
commonly-accepted maxim of animal experimentation is that animal use should be narrowly tailored
to answering questions which cannot be answered in other ways. Consider, for example, ``the 3Rs
alternatives'' approach to the use of animal experiments \cite{nrc.care}. In our case, we leverage
computer simulations for initial investigations into co-processor design, gathering evidence in
preparation for future \textit{in vivo} experiments. In the Discussion section, we explore
potential paths for translation of this work to \textit{in vivo} experiments.

A previous example of such a simulation approach is the work of Dura-Bernal et al.
\cite{bernal.sim}. In this work, the authors used a simulated spiking neural network
to train a stimulation agent. Their stimulation agent sought to restore the network's
control of a simulated arm to reach a target, after a simulated lesion was
applied. Similar to our approach, the authors simulated lesions by effectively removing parts
of their simulated network, or by removing connections between parts of the network.
As the authors point out, there exists only limited ability to probe a neural circuit
\textit{in vivo} in order to perform learning. As a result, we first need to design
our approach through the use of an admissible simulation. The key question then is:
how can we design an admissible simulation, i.e. one which provides a test bed on
which we can demonstrate key properties of our co-processor designs? We answer this question in our
work by adopting a a previously published model of cortical grasping \cite{michaels.mrnn} that has
been shown to replicate properties of the cortical circuit in the biological brain based on fits to
nonhuman primate data.

Through our simulation, we explored what properties of the co-processor allow successful
adaptation to the short-term dynamics of the cortical model as it is being stimulated,
as well as adaptation to longer-term connectivity changes in the cortical model.
We present a training method for neural co-processors for learning optimal stimulation
patterns that drive improvements in external task performance, while also adapting to the
non-stationarity of the stimulated neural circuits.

@@

\begin{figure}
    \centering
    \begin{subfigure}[c]{0.88\textwidth}
		\centering
		\includegraphics[width=\textwidth]{weill_arch.png}
		\caption{}
	\end{subfigure}
	\hfill
    \begin{subfigure}[c]{0.88\textwidth}
		\centering
		\includegraphics[width=\textwidth]{cpn_michaels_arch_labeled.eps}
		\caption{}
	\end{subfigure}
	\hfill
    \caption{{\bf Neural co-processor for restoring function after a brain injury}.
    An artificial neural network called the “Co-Processor Network” (CPN) is used to
    map input neural activity patterns in an area A to output stimulation patterns
    in same or other areas B in order to achieve a neural or behavioral goal using
    another ANN, an “Emulator Network” (EN) - see text for details. (a) The example
    here shows the CPN creating a new information processing pathway between prefrontal
    cortex and motor cortex, bypassing an intermediate area affected by brain injury
    (e.g., stroke). Adapted from \cite{rao.coproc}, (b) Our current study involves a
    simulated cortical grasping circuit. The CPN and EN both receive simulated brain
    recordings, which are created according to a recording model (see Methods).
    The CPN outputs stimulation parameters, which are applied to the simulated
    grasping circuit, according to a stimulation model. The EN models the relationship between
    the     stimulation, brain recordings, and external task.}
    \label{fig:arch}
\end{figure}

\subsection{Architecture Overview}
First, we present the architecture of our neural co-processor design. This design aims to solve two
fundamental challenges in using neural stimulation to improve external task performance.
First, to restore function for complex tasks (such as grasping), it is difficult to determine
the mapping between observed neural activity and the stimulation patterns to be applied to a downstream
motor area. As a result, the co-processor must learn what stimulation pattern is appropriate for achieving the
external task given the current neural activity. Unfortunately, in most cases, we do not know what the correct
stimulation pattern is for any given input activity pattern, a precondition for using supervised deep learning
methods for training the co-processor. Stimulation shapes the nonlinear dynamics of circuits in the brain
in complex ways, leading to complex effects in external behavior. It is therefore not obvious what
stimulation patterns will produce a desired behavior.

A neural co-processor attempts to solve these problems with a pair of artificial neural networks 
(Fig.~\ref{fig:arch}):
\begin{itemize}
    \item a ``Co-Processor Network'' (CPN), which is a recurrent neural network (RNN) which maps neural activity, and possibly
	      data from external sensors, to appropriate stimulation parameters.
    \item an ``Emulator Network'' (EN), also an RNN which models the effect of stimulation on neural dynamics and
          behavior for the external task.
\end{itemize}

The CPN can be trained using the backpropagation algorithm, the workhorse of deep learning for training ANNs.
However, backpropagation requires the error between the output of the CPN and a desired output, and as
discussed above, we do not have the desired output stimulation pattern. We do however know what the desired
output behavior in a task should be, e.g., a particular kind of grasp for a particular object or a particular
type of neural activity in a brain area corresponding to healthy activity. We can therefore compute the error
between this desired behavior and the actual behavior caused by the CPN due to stimulation. How do we
backpropagate this external task error to update the parameters of the CPN so that it learns to produce
stimulation patterns that are optimal for the task?

We use a trained EN to backpropagate the external task error to the CPN, which then uses this backpropagated
error to update its parameters. We train the EN to predict task-relevant parameters - a prediction of muscle
velocities in our case - given randomly sampled stimulation patterns or stimulation
patterns output by the CPN, and any measured neural activity. If the EN is trained to a sufficiently
high level of precision, it can be used as a function approximator for the subject's true ``stimulation function''.
When training the CPN, we treat the EN's output as the actual task output (e.g., actual grasp
behavior or muscle velocities). We backpropagate through the EN (without changing its parameters) and then the
CPN (changing its parameters) the error between the EN's output and the desired task output (see Fig. \ref{fig:arch}).

In our experiments, the EN was a single-layer fully-connected long short-term memory (LSTM) recurrent
neural network, with hyperbolic tangent ($tanh$) activations, and a linear readout. It had 87 LSTM neurons,
the number being chosen somewhat arbitrarily as a function of the input and output vector sizes. We found that
varying this neuron count did not drastically change results. Although other architectures could also be used,
we found that this LSTM architecture allows the EN to continuously adapt to long-running dependencies in the
simulated neural dynamics, far better than a vanilla RNN. The CPN had an almost identical architecture, but with
61 LSTM neurons. As with the CPN, this neuron count was chosen as a function of the input and output vector sizes,
and increasing it had little effect on results. There is no requirement for the EN and CPN to have similar network
architectures, but we found that these choices worked well in our experiments.

Note that the EN is more general than traditional models of neurostimulation which attempt to predict the effects of
stimulation on { \em neural activity}. The EN in a neural co-processor predicts the effects of stimulation (taking into
account ongoing neural dynamics) on {\em task performance}. This provides the key functionality needed to train the
CPN. In the special case where the task involves driving neural circuits in the brain to desired neural activities,
the EN reduces to more traditional models of stimulation.

For comparison, consider an EN architecture based on a traditional RNN with a nonlinearity. This is a nonlinear version
of the common linear time-invariant state space stimulation model studied in previous research (see, e.g., \cite{shanechi.stimmodel}). 
We found, however, that compared to an LSTM-based EN, the linear model and the vanilla nonlinear RNN are both not
sufficiently powerful to capture the long-term dependencies in stimulation effects needed for the CPN to learn well.
Our use of an LSTM for learning an EN builds on previous work using LSTMs for predicting the evolution of local field
potentials \cite{kim.lstm} and blood-oxygen-level dependent (BOLD) hemodynamic responses \cite{guclu.lstm} over time.

\subsection{Simulation Overview}
To test the feasibility of the neural co-processor approach, we used a previously published cortical model for
grasping in a nonhuman primate (NHP) brain. Using such a model allowed us to explore some of the critical
architectural choices and training algorithms for neural co-processors, enabling us to rapidly and cheaply iterate on our
design, laying the groundwork for future \textit{in vivo} experiments.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\usepackage{graphicx,epstopdf}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{lscape}
\usepackage{hyperref}
\graphicspath{ {./figs/} }
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\begin{document}

@@

% Authorship temporarily removed for double-anonymous review
%\author{Matthew J Bryan$^{1}$, Linxing Preston Jiang$^{1}$, Rajesh P N Rao$^{1}$}

%\address{$^{1}$ Neural Systems Laboratory, Paul G. Allen School of Computer
%Science \& Engineering, University of Washington, Box 352350,
%Seattle, WA 98105, USA}

%\ead{\{mmattb,prestonj,rao\}@cs.washington.edu}
\vspace{10pt}
\begin{indented}
\item[]November 2022
\end{indented}

\begin{abstract}
\textit{Objective} A major challenge in closed-loop brain-computer interfaces (BCIs)
is finding optimal stimulation patterns as a function of ongoing neural activity for
different subjects and for different objectives. Traditional approaches, such as
those currently used for deep brain stimulation (DBS), have largely followed a manual
trial-and-error strategy to search for effective open-loop stimulation parameters,
a strategy that is inefficient and does not generalize to closed-loop
activity-dependent stimulation.
\textit{Approach} To achieve goal-directed closed-loop neurostimulation, we propose
the use of brain co-processors, devices which exploit artificial intelligence (AI)
to shape neural activity and bridge injured neural circuits for targeted repair and
rehabilitation.  Here we investigate a specific type of co-processor called a ``neural
co-processor'' which uses artificial neural networks (ANNs) and deep learning to learn optimal
closed-loop stimulation policies. The co-processor adapts the stimulation policy as
the biological circuit itself adapts to the stimulation, achieving a form of brain-device
co-adaptation. We tested the neural co-processor's ability to restore function after
stroke by simulating a variety of lesions in a previously published cortical model
of grasping.  We additionally tested the ability of our co-processor
to adapt its stimulation as the simulated cortical network and simulated sensors
underwent changes.
\textit{Main results} Our results show that a neural co-processor can restore
reaching and grasping function after a simulated stroke in a cortical model, achieving
recovery towards healthy function in the range 75-90\%. Our co-processor successfully
co-adapted to accomplish the reach-and-grasp task after a variety of lesions to the
simulated cortical circuit.
\textit{Significance} Our results provide the first proof-of-concept demonstration,
using computer simulations, of a neural co-processor for activity-dependent closed-loop
neurosimulation for optimizing a rehabilitation goal after injury. Our results provide
insights on how such co-processors may eventually be developed for \textit{in vivo} use
to learn complex adaptive stimulation policies for a variety of neural rehabilitation
and neuroprosthetic applications. 
\end{abstract}

\vspace{2pc}
\noindent{\it Keywords}: brain-computer interface, neural co-processor, AI, machine learning, deep learning, neural networks, stimulation, computational models
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

@@

The co-processor framework also allows co-adaptation with biological circuits in the brain by updating its stimulation policy, while the brain updates its own response to the stimulation via adaptation and neural plasticity, or updates due to other reasons. This allows the co-processor to continually optimize its outputs for the desired optimization function in the presence of significant non-stationarity in the brain.

Here we use computer simulations to demonstrate a neural co-processor that restores
movement in a computational model of cortical networks controlling a limb, after a simulated stroke affects the ability to
use that limb. Our demonstration combines:
\begin{itemize}
	\item An emulation model based on ANNs, which models the relationship between 
	      stimulation, decoded brain activity, and task performance.
	\item An AI agent based on ANNs which determines the best stimulation to apply in a closed-loop fashion in real time.
\end{itemize}

\section{Background}
Significant advances have been made in understanding and modeling the effects of electrical stimulation
on the brain. Researchers have explored how information can be biomimetically or
artificially encoded and delivered via stimulation to neuronal networks in the brain and
other regions of the nervous system for auditory \cite{niparko.cochlear}, visual \cite{weiland.retinal},
proprioceptive \cite{tomlinson.propr}, and tactile
\cite{tabot.tact, tyler.tact, dadarlat.tact, sharlene.tact, cronin.tact} perception.
Advances have also been made in modeling the effects of stimulation over large scale, multi-region
networks, and across time \cite{shanechi.stimmodel}. Some models can additionally adapt to ongoing
changes in the brain, including changes due to the stimulation itself
\cite{tafazoli.acls}. For our simulations described below, we use a stimulation
model, not unlike those cited above, which seeks to account for both network dynamics
and non-stationarity. In addition to training the model to have a strong ability to predict
the effect of stimulation, we additionally adapt it to be useful for learning an
optimal stimulation policy, a property distinct from predictive
power alone.

Researchers have also explored both open- and closed-loop stimulation protocols for
treating a variety of disorders. Open loop stimulation has been effective in
treating Parkinson's Disease (PD) \cite{benabid.parkinsons}, as well as various
psychiatric disorders \cite{holtzheimer.psy, kisely.psy, fraint.psy}.
In research more directly related to our work, Khanna et al. \cite{khanna.openloop}, investigated 
the use of open loop stimulation in restoring dexterity after a lesion in a nonhuman primate's (NHP's) motor cortex.
The authors demonstrate that the use of low-frequency alternating current, applied epidurally,
can improve grasp performance.

While open loop stimulation techniques have yielded clinically useful results, results in other domains have been mixed, such as in visual
prostheses \cite{bosking.visual}, and in invoking somatosensory feedback
\cite{cronin.tact}. We believe this is due to the stimulation not being conditioned
on the ongoing dynamics of the neural circuit being stimulated. From moment to moment and
throughout the day, a neuronal circuit in the brain can be expected to respond differently even when the same stimulation parameters are used,
due to the multitude of different external and internal inputs influencing the circuit's ongoing activity. Stimulation
therefore needs to be proactively adapted in response. This need is even
greater over longer time scales as the effects of plasticity, changes in clinical conditions, and ageing change
the dynamics and connectivity of the brain. Closed-loop stimulation may also provide means
to better regulate the energy use of an implanted stimulator, allowing it to
intelligently regulate when to apply stimulation, in order to preserve implant
battery life. Another benefit is that closed-loop stimulation offers an opportunity to minimize the 
side-effects of stimulation, through real time regulation
of the stimulation parameters, such as in the use of deep brain stimulation (DBS) in
PD patients \cite{little.park}. In recent years, closed-loop stimulation has been used to aid in learning
new memories after some impairment \cite{berger.closedloop, kahana.biomarker},
to replay visually-invoked activations \cite{tafazoli.acls}, and for optogenetic
control of a thalamocortical circuit \cite{bolus.opto}, among others.

A major open question is: how does one leverage closed-loop control for real-time
co-adaptation with the brain to accomplish an external task? ``Co-adaptation'' here refers
to the ability of a BCI to adapt its stimulation regime to the ongoing
changes in the circuit it is stimulating, and to adapt with that circuit to accomplish
the external task (e.g., grasping). The neural co-processor we present here provides one
potential model for accomplishing that. Through the use of deep learning, a neural 
co-processor model co-adapts its AI, which governs the
stimulation, with the neural circuit being stimulated.

For a neurologically complex task such as grasping, it is not possible to design
\textit{a priori} a fixed real-time controller for the (potentially impaired) neural circuits involved in the task. That is
due in large part to the variability of circuits from subject-to-subject,
as well as variations in the placement of sensors and stimulators in the
brain. The only plausible path to implementing a real-time controller is therefore to allow the device to adapt to the subject,
long-term changes in their brain activities, and variability in the sensors, stimulators and hardware. Our proposed neural
co-processors seek to accomplish such adaptation through ANNs and deep learning.

To gain insights into neural co-processors before testing them in \textit{in vivo} experiments, we investigated a number of crucial design elements through the use of a neural network model of the cortical areas involved in 
grasping, presented previously by Michaels et al. \cite{michaels.mrnn}. We
explored what properties of the co-processor allow
successful adaptation to the short-term dynamics of the cortical model as it is being stimulated,
as well as adaptation to longer-term connectivity changes in the cortical model.
We present a training method for neural co-processors for learning optimal stimulation patterns that drive improvements in external task performance,
while also adapting to the non-stationarity of the brain.

@@

\begin{figure}
    \centering
    \begin{subfigure}[c]{0.75\textwidth}
		\centering
		\includegraphics[width=\textwidth]{weill_arch.png}
		\caption{}
	\end{subfigure}
	\hfill
    \begin{subfigure}[c]{0.75\textwidth}
		\centering
		\includegraphics[width=\textwidth]{cpn_michaels_arch_labeled.eps}
		\caption{}
	\end{subfigure}
	\hfill
    \caption{{\bf Neural co-processor for restoring function after a brain injury}.
    An artificial neural network called the “Co-Processor Network” (CPN) is used to
    map input neural activity patterns in an area A to output stimulation patterns
    in same or other areas B in order to achieve a neural or behavioral goal using
    another ANN, an “Emulator Network” (EN) - see text for details. (a) The example
    here shows the CPN creating a new information processing pathway between prefrontal
    cortex and motor cortex, bypassing an intermediate area affected by brain injury
    (e.g., stroke). Adapted from \cite{rao.coproc}, (b) Our current study involves a
    simulated cortical grasping circuit. The CPN and EN both receive simulated brain
    recordings, which are created according to a recording model (see Methods).
    The CPN outputs stimulation parameters, which are applied to the simulated
    grasping circuit, according to a stimulation model. The EN models the relationship between the
    stimulation, brain recordings, and external task.}
    \label{fig:arch}
\end{figure}

\subsection{Architecture Overview}
First, we present the architecture of our neural co-processor design. This design aims to solve two
fundamental challenges in using neural stimulation to improve external task performance.
First, to restore function for complex tasks, such as grasping, the mapping between
neural activity, such as the activity encoding the intention to grasp, and the stimulation patterns
to be applied to a downstream motor area to enable grasping cannot be pre-determined.
As a result, the co-processor must learn what stimulation pattern is appropriate for achieving the
external task. Unfortunately, in most cases, we do not know the optimal stimulation patterns to train
the co-processor. This is because stimulation shapes the nonlinear dynamics of circuits in the brain
in complex ways, leading to complex effects in external behavior. It is therefore not obvious what
stimulation patterns will produce a desired behavior.

A neural co-processor attempts to solve these problems with a pair of artificial neural networks (ANNs)
(Fig.~\ref{fig:arch}):
\begin{itemize}
    \item a ``Co-Processor Network'' (CPN), which maps neural activity, and possibly
	      data from external sensors, to appropriate stimulation parameters.
    \item an ``Emulator Network'' (EN) which models the effect of stimulation on neural dynamics and
          behavior for the external task.
\end{itemize}

The CPN can be trained using the backpropagation algorithm, the workhorse of deep learning for training ANNs.
However, backpropagation requires the error between the output of the CPN and a desired output, and as
discussed above, we do not have the desired output stimulation pattern. We do however know what the desired
output behavior in a task should be, e.g., a particular kind of grasp for a particular object or a particular
type of neural activity in a brain area corresponding to healthy activity. We can therefore compute the error
between this desired behavior and the actual behavior caused by the CPN due to stimulation. How do we
backpropagate this error, which is external to the CPN, to update the parameters of the CPN?

We use a trained EN to backpropagate the external task error to the CPN. We train the EN to predict
task-relevant parameters - a prediction of muscle velocities in our case - given the stimulation
parameters output by the CPN and any measured neural activity. If the EN is trained to a sufficiently
high level of precision, it can be used as a function approximator mapping stimulation to task output.
When training the CPN, we treat the EN's output as the actual task output (e.g., actual grasp
parameters).  We backpropagate through the EN (without changing its weights) the error between
the EN's output and the desired task output and use this backpropagated error to train the
CPN (see Fig. \ref{fig:arch}).

In our experiments, the EN was a single layer fully connected long short-term memory (LSTM) recurrent
neural network (RNN), with hyperbolic tangent ($tanh$) activations, and a linear readout. It has 87 LSTM neurons,
which was chosen somewhat arbitrarily as a function of the input and output vector sizes. We found varying this neuron
count did not drastically change results. Although other architectures could also be used, we found that this
LSTM architecture allows the EN to continuously adapt to long-running dependencies in the simulated neural dynamics, 
ar better than a vanilla RNN. The CPN had an almost identical architecture, but with 61 LSTM neurons. As with the CPN,
this neuron count was chosen as a function of the input and output vector sizes, but varying it upward had little effect on results.
There is no requirement for the EN and CPN to have similar network architectures, but we found that these choices worked
well in our experiments.

Note that the EN is more general than traditional models of neurostimulation which attempt to predict the effects of
stimulation on \em{neural activity}. The EN in a neural co-processor predicts the effects of stimulation (taking into
account ongoing neural dynamics) on \em{task performance}. This provides the key functionality needed to train the
CPN. In the special case where the task involves driving neural circuits in the brain to desired neural activities,
the EN reduces to more traditional models of stimulation.

For comparison, consider an EN architecture based on a traditional RNN with a nonlinearity. This is a nonlinear version
of the common linear time-invariant state space model of stimulation (see, e.g.,  \cite{shanechi.stimmodel}). 
We found, however, that compared to LSTM-based EN, the linear model and the traditional nonlinear RNN are both not
sufficiently powerful to capture the long-term dependencies in stimulation effects needed for the CPN to learn well.
Our use of an LSTM-based EN builds on previous work modeling biological neural networks for predicting local field potentials
\cite{kim.lstm}, and modeling stimulation \cite{guclu.lstm}.

\subsection{Simulation Overview}
To test the feasibility of the neural co-processor approach, we used a previously published model of a
grasping circuit in a nonhuman primate (NHP) brain. Using such a model allows us to explore some of the critical
architectural choices and training algorithms, enabling us to rapidly and cheaply iterate on our
design, prior to any \textit{in vivo} experiments.

In the present study, we use the grasping network model of Michaels et al. \cite{michaels.mrnn}
(Fig.~\ref{fig:michaels}a). The network, representing multiple interconnected cortical areas, was trained
to mimic the grasping circuits of NHP subjects engaged in a delayed reach-to-grasp task.
The model's design draws on a body of literature focused on architectures and training methods for RNNs which
seek to create artificial neural networks with activation dynamics similar to biological circuits,
including circuits for delayed grasping tasks \cite{susillo.mrnn}. The model consists of three
``modular'' vanilla RNNs (mRNNs), representing the cortical areas AIP, F5 and M1 respectively, and a linear
readout layer (Fig.~\ref{fig:michaels}a). Each ``module'' consists of 100 vanilla
RNN neurons, with a nonlinearity applied on the outputs. The modules are internally fully connected,
and are connected to each other sparsely (10\% connectivity in our case). The inputs to the network are visual features
representing the object to be grasped, as well as a hold signal. The visual features
intend to capture the features represented in the subject's visual cortex. They were extracted using
VGGNet \cite{simonyan.vgg} from 3D renderings of the same objects which the subjects grasped. The hold signal is
a boolean which encodes the point in the experiment when the subject began to reach. The outputs of the
network are muscle length velocities for the shoulder, arm, and hand of the subject. The actual
velocities were captured with a motion capture glove, and the Michaels et al. network model 
was trained to recapitulate those grasping motions.\footnote{Data and trained models from this work
were supplied to us by the lead author of \cite{michaels.mrnn}. We re-implemented their model in PyTorch,
and used their trained parameters for an arbitrarily chosen subject.}
